---
title: "Unicef's Mission to democratising Sanitation"
subtitle: "Unicef's Report on people using Limited Sanitation Services"
author: "Gemma Cox"
date: "2024-04-25"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(maps)
library(plotly)
library(ggplot2)

#data
unicef_metadata <- read_csv("unicef_metadata.csv")
unicef_indicator_1 <- read_csv("unicef_indicator_1.csv")
data_right_3 <- read_csv("data_right_3.csv")

# exercise 1
data_join <- full_join(unicef_metadata, unicef_indicator_1)
data_join <- full_join(unicef_metadata, unicef_indicator_1, 
                       by = c("country", "year" = "time_period"))

# exercise 3
data_join <- full_join(unicef_metadata, data_right_3)
data_join <- full_join(unicef_metadata, data_right_3, by = c("country"))

# final data object
data_join <- unicef_metadata %>%
  full_join(unicef_indicator_1, 
            by = c("country", "year" = "time_period")) %>%
  full_join(data_right_3, by = c("country"))
```
As the life expectancy age increases throughout the world, there are still parts
of the world that are left behind.

## World Map
```{r}
#| fig-cap: "As you can see, developed countries have a significantly higher 
#| life expectancy than developing countries, particularly in parts of _Africa_, 
#| where the life expectancy can go as low as 60 years old."

map_world <- map_data("world")

map_data_join_2018 <- data_join %>%
  filter(year == 2018) %>%
  full_join(map_world, by = c("country" = "region"))

j <- ggplot(map_data_join_2018) + 
  aes(x = long, y = lat, group = group, fill = `Life expectancy at birth, total (years)`) + 
  geom_polygon() + 
  labs(x = "", y = "", title = "Map revealing the life expectancy at birth in the year 2018") + 
  theme_classic() +
  theme(text = element_text(family = "helvetica")) + 
  scale_fill_gradient(low = "red", high = "green", na.value = "grey", name = "Life Expectancy")

# Convert ggplot object to plotly object
ggplotly(j)
```

## Bar Chart
```{r}
#| fig-cap: "Today we look into the life expectancy figures, as well as observe
#| the change in % of people who have limited sanitation services (i.e. shared
#| use of toilet facilities). Between 2016-2018, a staggering amount of the
#| population of _Africa_, _Asia_ and the _Americas_ are living in these poor 
#| conditions."

data_join_bar <- full_join(data_right_3, unicef_indicator_1, by = c("country"))

k <- ggplot(subset(data_join_bar, time_period %in% c("2016", "2017", "2018")), 
       aes(x=continent, y=obs_value, fill = continent)) + 
  geom_bar(stat="identity", width=0.5) + 
  facet_wrap(~ time_period, nrow = 1) + 
  scale_x_discrete(limits = c("Africa", "Europe", "Asia", "Americas", "Oceania")) + 
  scale_y_continuous(limits = c(0,100)) + 
  labs(x = "", y = "People who have limited sanitation", title = 
  "The Percentage of people who have limited sanitation services between 2016-18",
  size = 0.8) + theme_classic() +
  theme(text = element_text(family = "helvetica")) + scale_fill_manual(values = 
  c("#00AFBB", "#4FB05E", "#DB245E", "#FFE800", "#FFC0CB"))

# Convert ggplot object to plotly object
ggplotly(k)

```

## Timeseries Plot
```{r}
#| fig-cap: "Interestingly, this timeseries shows us that over the last 20
#| years, _Ghana_ and _American_ _Samoa_ have the most people using limited
#| sanitation services, however, both of these outliers have gradually improved."

timeseries_plot_1 <- full_join(data_right_3, unicef_indicator_1, by = c("country"))

p <- ggplot(timeseries_plot_1) + 
  aes(x = time_period, y = obs_value, group = country, color = continent) + 
  geom_line(size = 1.4) + 
  scale_color_manual(values = c("#00AFBB", "#4FB05E", "#DB245E", "#FFE800", "#FFC0CB")) + 
  scale_y_continuous(limits = c(0, 80)) + 
  theme_minimal() + 
  labs(x = "year", y = "People who have limited sanitation",
       title = "The Percentage of People who have limited sanitation services") + 
  theme_classic() + 
  theme(text = element_text(family = "helvetica"))

# Convert ggplot object to plotly object
ggplotly(p)
```
__Question__: is there a notable correlation between life
expectancy correlate with the % of people who have limited sanitation services?

## Scatter Plot
```{r}
#| fig-cap: "We can identify that the majority of _African_ _countries_ that 
#| have a __high__ __life__ __expectancy__, have a low % of the people 
#| who have limited sanitation services."

scatter_plot_filtered <- unicef_metadata %>%
  full_join(unicef_indicator_1, by = c("country", "year" = "time_period")) %>%
  full_join(data_right_3, by = c("country")) %>%
  filter(continent == "Africa", year %in% c("2016", "2017", "2018"))

# Create ggplot object with regression line
q <- ggplot(scatter_plot_filtered) + 
  aes(x = obs_value, 
      y = `Life expectancy at birth, total (years)`, 
      color = as.factor(year)) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm", se = FALSE) +  # Add regression line
  labs(x = "Percentage of people with limited sanitation services", 
       y = "Life Expectancy at birth, total (years)", 
       title = "Evolution of the relationship between Life expectancy and people who have limited sanitation services in _Africa_ (2016-2018)",
       color = "Year") +
  theme_classic() + 
  theme(text = element_text(family = "helvetica"))

# Convert ggplot object to plotly object
ggplotly(q)
```
## Conclusion
In conclusion, this negative correlation proves that both low life expectancy
and high percentage of pop using limited sanitation services correlate. In order
for us to be on target with our Goal 6: Clean water and sanitation for all for
2030, UNICEF are working across the sanitation management chain, aiming to
reduce high-risk hygiene practices, particularly for those who live in
high-density urban areas.

To find out more,
visit [UNICEF](https://www.unicef.org/esa/sanitation-and-hygiene)
